---
title: "Decision Trees"
author: "INSOFE Lab Activity on Decision Trees Abhishek Gautam"
date: "23 July 2017"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
---

# C5.0 Trees

**NOTE** Before starting this assignment please remember to clear your environment, you can do that by running the following code chunk

```{r}
rm(list = ls(all = TRUE))

```

## Goal

* The goal of this activity is to predict wether a patient has liver disease or not based on various patient related attributes


## Agenda 

* Get the data

* Data Pre-processing

* Build a model

* Predictions

* Communication


## Reading & Understanding the Data

### Read the Data

Make sure the dataset is located in your current working directory, or else you can change your working directory using the "setwd()" function.

```{r}
disease_data = read.csv(file = "ilpd_data.csv", header = TRUE, sep = ",")

```

### Understand the data

* Use the str(), summary(), head() and tail() functions to get the dimensions and types of attributes in the dataset

* The dataset has 582 observations and 11 variables

- __The variable descriptions are given below:__

1 - age	 : Age of the patient

2 - gender : Gender of the patient

3 - TB : Total Bilirubin content

4 - DB : Direct Bilirubin content

5 - alk_phos : Alkaline Phosphotase content

6 - alamine : Alamine Aminotransferase content

7 - aspartate : Aspartate Aminotransferase content

8 - TP : Total Protiens content

9 - albumin : 	Albumin content

10 - A/G : Ratio of Albumin and Globulin 

11 - Disease : Whether the patient has liver disease or not 


```{r}
str(disease_data)
summary(disease_data)
```

```{r}
head(disease_data, n = 10)
tail(disease_data, n=10)


```


## Data Pre-processing

### Verify Data Integrity

* Verify if the dataset has missing values

```{r}
sum(is.na(x = disease_data))
sum(is.na(x = disease_data$disease))

```

* Verify the data types assigned to the variables in the dataset

```{r}

str(disease_data)

```

### Split the Data into train and test sets

* Use stratified sampling to split the data into train/test sets (70/30)

* Use the createDataPartition() function from the caret package to do stratified sampling

```{r}
library(caret)
set.seed(9890)
train_rows = createDataPartition(y = disease_data$disease, p = 0.7, list = FALSE)

train_data = disease_data[train_rows,]
test_data = disease_data[-train_rows,]

```

### Impute the missing values

* Impute missing values using knnImputation() function in both the train and test datasets

```{r}
library(DMwR)
train_data_KNN = knnImputation(data = train_data,k = 7)
test_data_KNN = knnImputation(data = test_data,k = 7)
sum(is.na(test_data_KNN))
sum(is.na(train_data_KNN))

```

## Build a  Decision Tree

### Model the tree

* Use Quinlan's C5.0 decision tree algorithm implementation from the C50 package to build your decision tree

```{r}
library(C50)

tree_c50 = C5.0(disease~. , train_data_KNN)
tree_c50

```

* Build a rules based tree

```{r}

rules_c50 = C5.0(disease~. ,train_data, rules = TRUE)

```

### Variable Importance in trees

* Find the importance of each variable in the dataset

```{r}
?C5imp
C5imp(tree_c50, metric = "usage")


```

### Rules from trees

* Understand the summary of the returned c5.0 rules based on the decision tree model


```{r}

summary(rules_c50)

```


### Plotting the tree

* Call the plot function on the tree object to visualize the tree

```{r, fig.width= 35, fig.height=15}

plot(tree_c50)

```


## Evaluating the model

### Predictions on the test data

* Evaluate the decision tree using the standard error metrics on test data

```{r}

prediction = predict(tree_c50, test_data)

```

* Report error metrics for classification on test data

```{r}

library(caret)

confusionMatrix(prediction, test_data$disease)

```

# CART Trees

**NOTE** Before starting this assignment please remember to clear your environment, you can do that by running the following code chunk

```{r}

#rm(list = ls(all  =TRUE))

```

* The classification and regression trees use gini index in place of the gain ratio (based on information gain) used by the ID3 based algorithms, such as c4.5 and c5.0

## Goal

* The goal of this activity is to predict the heating load of a residential building, if the building parameters are given

* Hence, in the future architects would be able to build more energy efficient buildings as they can optimize the building parameters to reduce the heating load

## Agenda 

* Get the data

* Data Pre-processing

* Build a model

* Predictions

* Communication


## Reading & Understanding the Data

### Read the Data

* Make sure the dataset is located in your current working directory, or else you can change your working directory using the "setwd()" function.

```{r}

building_energy_data = read.csv(file = "building_energy.csv", header = TRUE, sep = ",")

```

### Understand the data

* Use the str(), summary(), head() and tail() functions to get the dimensions and types of attributes in the dataset

* The dataset has 768 observations and 9 variables

```{r}
str(building_energy_data)
summary(building_energy_data)

head(building_energy_data, n = 10)
tail(building_energy_data, n=10)


```

* The variable names are self explanatory, for further information visit http://www.sciencedirect.com/science/article/pii/S037877881200151X

## Data Pre-processing

### Verify Data Integrity

* Verify if the dataset has missing values

```{r}

sum(is.na(building_energy_data))

```

* Verify the data types assigned to the variables in the dataset

```{r}
str(building_energy_data)
# Enter answer here
#Relative Compactness  =NUMBER
#Surface area          =NUMBER
#Wall area             =NUMBER
#Roof area             =NUMBER
#Overall height        =NUMBER
#Orientation           =Integer
#Glazing area          =NUMBER
#Glazing area distribution =NUMBER
#Heating load          =NUMBER


```

### Split the Data

* Split the data into train/test sets (70/30)

```{r}

library(caret)
set.seed(123)
train_rows = createDataPartition(y = building_energy_data$heating_load, p = 0.7, list = FALSE)

train_data = building_energy_data[train_rows,]
test_data = building_energy_data[-train_rows,]

```

## Build a Regression Tree

### Model the tree

* Use the rpart package to build a cart tree to predict the heating load

```{r}

library(rpart)

tree = rpart(formula = heating_load~., data = train_data)


```

### Tree Explicability

* Print the variable importance

```{r}

printcp(tree)

```

* Plot the regression tree

```{r, fig.width=8, fig.height=5}

library(rpart.plot)

rpart.plot(tree)


```

## Evaluation on Test Data

* Report error metrics on the test data

```{r}
prediction = predict(tree, test_data)

```

```{r}

library(DMwR)

regr.eval(test_data$heating_load, prediction)
```




















